---
title: "DE Analysis"
author: "Christof Seiler"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will follow chapter 4 from [here](https://bioconductor.org/books/3.18/OSCA.multisample/multi-sample-comparisons.html).

## Packages

Install the necessary R package from Bioconductor.

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("MouseGastrulationData", "scater", "scran", "batchelor"))
```

## Data

Load data going through the usual preprocessing steps from chapter 10.

```{r}
#--- loading ---#
library(MouseGastrulationData)
sce.chimera <- WTChimeraData(samples=5:10)
counts(sce.chimera) <- as(counts(sce.chimera), "CsparseMatrix")
sce.chimera

#--- feature-annotation ---#
library(scater)
rownames(sce.chimera) <- uniquifyFeatureNames(
    rowData(sce.chimera)$ENSEMBL, rowData(sce.chimera)$SYMBOL)

#--- quality-control ---#
drop <- sce.chimera$celltype.mapped %in% c("stripped", "Doublet")
sce.chimera <- sce.chimera[,!drop]

#--- normalization ---#
sce.chimera <- logNormCounts(sce.chimera)

#--- variance-modelling ---#
library(scran)
dec.chimera <- modelGeneVar(sce.chimera, block=sce.chimera$sample)
chosen.hvgs <- dec.chimera$bio > 0

#--- merging ---#
library(batchelor)
set.seed(01001001)
merged <- correctExperiments(sce.chimera, 
    batch=sce.chimera$sample, 
    subset.row=chosen.hvgs,
    PARAM=FastMnnParam(
        merge.order=list(
            list(1,3,5), # WT (3 replicates)
            list(2,4,6)  # td-Tomato (3 replicates)
        )
    )
)

#--- clustering ---#
g <- buildSNNGraph(merged, use.dimred="corrected")
clusters <- igraph::cluster_louvain(g)
colLabels(merged) <- factor(clusters$membership)

#--- dimensionality-reduction ---#
merged <- runTSNE(merged, dimred="corrected", external_neighbors=TRUE)
merged <- runUMAP(merged, dimred="corrected", external_neighbors=TRUE)
```

The t-SNE plot show the condition (tomato vs control) and the batch (pool 3, 4, and 5). The number of the left are the median center of each cell cluster.

```{r}
gridExtra::grid.arrange(
    plotTSNE(merged, colour_by="tomato", text_by="label"),
    plotTSNE(merged, colour_by=data.frame(pool=factor(merged$pool))),
    ncol=2
)
```

Compare the clusters with an larger cell atlas. Many clusters map to many cell types and the other way around. The color scale represents the log2-count for each combination.

```{r}
by.label <- table(colLabels(merged), merged$celltype.mapped)
pheatmap::pheatmap(log2(by.label+1), color=viridis::viridis(101))
```

## Pseudo-Bulk Analysis

Sum of cells per sample and cell type.

```{r}
# Using 'label' and 'sample' as our two factors; each column of the output
# corresponds to one unique combination of these two factors.
summed <- aggregateAcrossCells(merged, 
    id=colData(merged)[,c("celltype.mapped", "sample")])
summed
```

Differential analysis using `edgeR` for cell type Mesenchyme.

```{r}
label <- "Mesenchyme"
current <- summed[,label==summed$celltype.mapped]

# Creating up a DGEList object for use in edgeR:
library(edgeR)
y <- DGEList(counts(current), samples=colData(current))
y
```

Remove samples with very low library sizes.

```{r}
discarded <- current$ncells < 10
y <- y[,!discarded]
summary(discarded)
```

Remove low expressed genes.

```{r}
keep <- filterByExpr(y, group=current$tomato)
y <- y[keep,]
summary(keep)
```

Correct for composition biases.

```{r}
y <- calcNormFactors(y)
y$samples
```

Standard RNA-seq DE diagnostic plots.

```{r}
for (i in seq_len(ncol(y))) {
    plotMD(y, column=i)
}
```

MDS plot colored by condition.

```{r}
plotMDS(cpm(y, log=TRUE), 
    col=ifelse(y$samples$tomato, "red", "blue"))
```

Design matrix.

```{r}
design <- model.matrix(~factor(pool) + factor(tomato), y$samples)
design
```

Estimate dispersions.

```{r}
y <- estimateDisp(y, design)
summary(y$trended.dispersion)
plotBCV(y)
```

Estimate the quasi-likelihood dispersions.

```{r}
fit <- glmQLFit(y, design, robust=TRUE)
summary(fit$var.prior)
plotQLDisp(fit)
```

Differential tests.

```{r}
res <- glmQLFTest(fit, coef=ncol(design))
summary(decideTests(res))
topTags(res)
```

Very few genes are significant in this case.

Repeat the same procedure as above with one function for all cell types.

```{r}
# Removing all pseudo-bulk samples with 'insufficient' cells.
summed.filt <- summed[,summed$ncells >= 10]

de.results <- pseudoBulkDGE(summed.filt, 
    label=summed.filt$celltype.mapped,
    design=~factor(pool) + tomato,
    coef="tomatoTRUE",
    condition=summed.filt$tomato 
)

cur.results <- de.results[["Mesenchyme"]]
cur.results[order(cur.results$PValue),]
```

## Reproducibility

```{r}
sessionInfo()
```
