---
title: "Mixed Effect DE Analysis"
author: "Christof Seiler"
output: html_document
date: "`r gsub(' 0', ' ', format(Sys.time(), '%B %d, %Y'))`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will follow the `muscat` vignette on Bioconductor [here](https://bioconductor.org/packages/release/bioc/vignettes/muscat/inst/doc/analysis.html#differential-state-ds-analysis).

## Packages

Install the necessary R package from Bioconductor.

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("muscat", "scater", "UpSetR"))
```

## Data

Kang 2018 from the vignette.

```{r}
library(ExperimentHub)
eh <- ExperimentHub()
query(eh, "Kang")
(sce <- eh[["EH2259"]])
```

Preprocesssing.

```{r}
# remove undetected genes
sce <- sce[rowSums(counts(sce) > 0) > 0, ]
dim(sce)

# calculate per-cell quality control (QC) metrics
library(scater)
qc <- perCellQCMetrics(sce)

# remove cells with few or many detected genes
ol <- isOutlier(metric = qc$detected, nmads = 2, log = TRUE)
sce <- sce[, !ol]
dim(sce)

# remove lowly expressed genes
sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
dim(sce)

# compute sum-factors & normalize
sce <- computeLibraryFactors(sce)
sce <- logNormCounts(sce)
```

Data preparation.

```{r}
library(muscat)
sce$id <- paste0(sce$stim, sce$ind)
(sce <- prepSCE(sce, 
    kid = "cell", # subpopulation assignments
    gid = "stim",  # group IDs (ctrl/stim)
    sid = "id",   # sample IDs (ctrl/stim.1234)
    drop = TRUE))  # drop all other colData columns
```

Cluster-sample sizes.

```{r}
# nb. of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))
```

Dimension reduction.

```{r}
# wrapper to prettify reduced dimension plots
.plot_dr <- function(sce, dr, col)
  plotReducedDim(sce, dimred = dr, colour_by = col) +
    guides(fill = guide_legend(override.aes = list(alpha = 1, size = 3))) +
    theme_minimal() + theme(aspect.ratio = 1)

# downsample to max. 100 cells per cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
cs100 <- unlist(sapply(cs_by_k, function(u) 
  sample(u, min(length(u), 100))))

# plot t-SNE & UMAP colored by cluster & group ID
.plot_dr(sce[, cs100], "TSNE", "cluster_id")
.plot_dr(sce[, cs100], "TSNE", "group_id")
```

## Pseudobulk Analysis

Aggregate data.

```{r}
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id"))
# one sheet per subpopulation
assayNames(pb)
t(head(assay(pb)))
```

MDS plot.

```{r}
pb_mds <- pbMDS(pb)
```

Pseudobulk.

```{r}
# run DS analysis
res <- pbDS(pb, verbose = FALSE)
# access results table for 1st comparison
tbl <- res$table[[1]]
# one data.frame per cluster
names(tbl)
# view results for 1st cluster
k1 <- tbl[[1]]
head(format(k1[, -ncol(k1)], digits = 2))
```

Print results.

```{r}
library(dplyr)
# filter FDR < 5%, abs(logFC) > 1 & sort by adj. p-value
tbl_fil <- lapply(tbl, function(u) {
  u <- dplyr::filter(u, p_adj.loc < 0.05, abs(logFC) > 1)
  dplyr::arrange(u, p_adj.loc)
})

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil, nrow, numeric(1))
p_de <- format(n_de / nrow(sce) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)

# view top 2 hits in each cluster
top2 <- bind_rows(lapply(tbl_fil, top_n, 2, p_adj.loc))
format(top2[, -ncol(top2)], digits = 2)
```

Calculating expression frequencies.

```{r}
frq <- calcExprFreqs(sce, assay = "counts", th = 0)
# one sheet per cluster
assayNames(frq)
# expression frequencies in each
# sample & group; 1st cluster
t(head(assay(frq), 5))
```

Visualizing between-cluster concordance.

```{r}
library(purrr)
library(UpSetR)
de_gs_by_k <- map(tbl_fil, "gene")
upset(fromList(de_gs_by_k))
```

## Mixed Effect Analysis

Mixed models.

```{r}
library(BiocParallel)

if(file.exists("mm_dream.RData")) {
  
  load("mm_dream.RData")
  
} else {
  
  # 1st approach (takes about 40 min with 16 cores and 128 GB RAM)
  mm_dream <- mmDS(sce, method = "dream",
    n_cells = 10, n_samples = 2,
    min_count = 1, min_cells = 20,
    BPPARAM = MulticoreParam(workers = 16, progressbar = TRUE)
    )
  save(mm_dream, file = "mm_dream.RData")
  
}
```

Some other mixed models.

```{r eval=FALSE}
# 2nd & 3rd approach
mm_vst <- mmDS(sce, method = "vst", vst = "sctransform",
  BPPARAM = MulticoreParam(workers = 16, progressbar = TRUE))
save(mm_vst, file = "mm_vst.RData")
mm_nbinom <- mmDS(sce, method = "nbinom", 
  BPPARAM = MulticoreParam(workers = 16, progressbar = TRUE))
save(mm_nbinom, file = "mm_nbinom.RData")
```

## Comparison of Pseudobulk and Mixed Models

Scatter and 2d histograms comparing $p$-values form both models look very different.

```{r}
mixed <- mm_dream |> 
  bind_rows() |> 
  select(gene, cluster_id, p_val_mixed = p_val, p_adj_mixed = p_adj.glb)

bulk <- tbl |> 
  bind_rows() |> 
  select(gene, cluster_id, p_val_bulk = p_val, p_adj_bulk = p_adj.glb)

combo <- inner_join(mixed, bulk, by = c("gene", "cluster_id"))

ggplot(combo, aes(p_val_bulk, p_val_mixed)) + 
  geom_point() + 
  facet_wrap(~cluster_id, nrow = 2) +
  coord_fixed()

ggplot(combo, aes(p_val_bulk, p_val_mixed)) +
  geom_hex(bins = 30, aes(fill = stat(log(count)))) +
  facet_wrap(~cluster_id, nrow = 2) + 
  coord_fixed() +
  xlab("bulk p-values") + 
  ylab("mixed p-values")
```

What is the agreement between methods on interesting genes?

```{r}
alpha <- 0.05
combo |> 
  group_by(cluster_id) |>
  summarize(
  both_yes = mean(p_adj_mixed <= alpha & p_adj_bulk <= alpha),
  both_no  = mean(p_adj_mixed > alpha & p_adj_bulk > alpha),
  mixed_yes_bulk_no = mean(p_adj_mixed <= alpha & p_adj_bulk > alpha),
  mixed_no_bulk_yes = mean(p_adj_mixed > alpha & p_adj_bulk <= alpha)
  )
```

## Reproducibility

```{r}
sessionInfo()
```
