---
title: "Nonparametric DE Analysis"
author: "Christof Seiler"
output: html_document
date: "`r gsub(' 0', ' ', format(Sys.time(), '%B %d, %Y'))`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Setup nonparametric workflow. We will follow the `distinct` vignette on Bioconductor [here](https://bioconductor.org/packages/release/bioc/vignettes/distinct/inst/doc/distinct.html).

## Packages

Install the necessary R package from Bioconductor.

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# distinct fails to install with R 4.3.1 throwing error
# RcppParallel.h: No such file or directory
# it works with R 4.4.0

# BiocManager::install("distinct")

# use the GitHub version until ScienceApps update to R 4.4.*
devtools::install_github("https://github.com/SimoneTiberi/distinct")
```

Load packages.

```{r}
library(SingleCellExperiment)
library(distinct)
```

## Data

Load the Kang 18 dataset from their vignette.

```{r}
data("Kang_subset", package = "distinct")
Kang_subset
colData(Kang_subset)
Kang_subset@metadata$experiment_info
```

## Nonparametric Analysis

Create design matrix.

```{r}
samples = Kang_subset@metadata$experiment_info$sample_id
group = Kang_subset@metadata$experiment_info$stim
design = model.matrix(~group)
# rownames of the design must indicate sample ids:
rownames(design) = samples
design
```

Permutation test.

```{r}
set.seed(61217)
res = distinct_test(x = Kang_subset, 
                    name_assays_expression = "logcounts",
                    name_cluster = "cell",
                    name_sample = "sample_id",
                    design = design,
                    column_to_test = 2,
                    min_non_zero_cells = 20,
                    n_cores = 2)
head(res)
```

Compute log2-fold changes.

```{r}
res = log2_FC(res = res,
              x = Kang_subset, 
              name_assays_expression = "cpm",
              name_group = "stim",
              name_cluster = "cell")
head(res)
```

Print top overall hits.

```{r}
head(top_results(res))
```

Print cluster specific hits.

```{r}
top_results(res, cluster = "Dendritic cells")
```

Plot densities.

```{r}
plot_densities(x = Kang_subset,
               gene = "ISG15",
               cluster = "Dendritic cells",
               name_assays_expression = "logcounts",
               name_cluster = "cell",
               name_sample = "sample_id",
               name_group = "stim")
```

Plot average densities.

```{r}
plot_densities(x = Kang_subset,
               gene = "ISG15",
               cluster = "Dendritic cells",
               name_assays_expression = "logcounts",
               name_cluster = "cell",
               name_sample = "sample_id",
               name_group = "stim",
               group_level = TRUE)
```

## Reproducibility

```{r}
sessionInfo()
```
